---
title: "bikes"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Load and examine data. Bike usage versus hour of day. 
```{r}

library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)


bike_data = read.csv("london_merged.csv")

head(bike_data)
dim(bike_data)

#add time as a variable
bike_data = bike_data %>%
   mutate(time = as.numeric(substring(timestamp, 12, 13)))

#plot usage vs hour of day
hours = aggregate(bike_data, by=list(Time = bike_data$time), mean)
head(hours)

ggplot(hours, aes(x=time, y=cnt)) + geom_point()

#plot usage throughout all the days 

daily_bike_data = bike_data %>% mutate(day = as.Date(timestamp, format="%Y-%m-%d")) %>%
  group_by(day) %>% # group by the day column
  summarise(day_cnt=sum(cnt))  # calculate the SUM of all cnts that occurred on each day

dim(daily_bike_data)
head(daily_bike_data, 50)

#plot bike usage vs day
ggplot(daily_bike_data, aes(x = day, y = day_cnt)) + geom_point() + labs(title = "Total bike usage per day over the full 2 years")
```


## Usage versus Weather Code
```{r}
no_night <- bike_data %>%
  filter(time >= 4 & time <= 21)
boxplot(cnt~weather_code,
data= no_night,
main="Bike Count according to Weather Code",
xlab="Weather Code",
ylab="Bike Count",
col="yellow",
border= "blue"
)

```



## Linear Models of All Rides
```{r}
library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)

bike_data <- read.csv("london_merged.csv")
head(bike_data, 24)
dim(bike_data)

bike_data <- bike_data %>%
  mutate(time = as.numeric(substring(timestamp, 12, 13)))

model1 <- lm(cnt ~ hum, data = bike_data)
model2 <- lm(cnt ~ t1, data = bike_data)
model3 <- lm(cnt ~ t1 + hum + is_weekend, data = bike_data)

summary(model1)
summary(model2)
summary(model3)

mean(model1$residuals^2)
mean(model2$residuals^2)
mean(model3$residuals^2)

ggplot(bike_data, aes(y = cnt, x = t1, color = hum)) + geom_point()
```

## Linear Models of Weekend Rides
```{r}
weekend_bikes <- bike_data %>%
  filter(is_weekend == 1)%>%
  filter(time >= 4 & time <= 21)

model1_2 <- lm(cnt ~ hum, data = weekend_bikes)
model2_2 <- lm(cnt ~ t1, data = weekend_bikes)
model3_2 <- lm(cnt ~ t1 + hum, data = weekend_bikes)

mean(model1_2$residuals^2)
mean(model2_2$residuals^2)
mean(model3_2$residuals^2)

summary(model1_2)
summary(model2_2)
summary(model3_2)

ggplot(weekend_bikes, aes(y = cnt, x = t1, color = hum)) + geom_point()
```
## Temperature (how it feels)
```{r}


model3_3 <- lm(cnt ~ t2 + hum, data = weekend_bikes)

mean(model3_3$residuals^2)

summary(model3_3)

ggplot(weekend_bikes, aes(y = cnt, x = t2, color = hum)) + geom_point()

modt2_results <- data.frame(observed = weekend_bikes$cnt, predicted = model3_3$fitted.values, residual = model3_3$residuals)

ggplot(modt2_results, aes(y = observed, x = predicted))+ geom_point()
ggplot(modt2_results, aes(y = residual, x = predicted))+ geom_point() + geom_hline(yintercept = 0)

ggplot(modt2_results, aes(sample = residual)) + 
    geom_qq()
```

## Model Assumptions for Weekend Rides
```{r}
mod_results <- data.frame(observed = weekend_bikes$cnt, predicted = model3_2$fitted.values, residual = model3_2$residuals)

ggplot(mod_results, aes(y = observed, x = predicted))+ geom_point()
ggplot(mod_results, aes(y = residual, x = predicted))+ geom_point() + geom_hline(yintercept = 0)

ggplot(mod_results, aes(sample = residual)) + 
    geom_qq()
```


## Determining whether current weather or previous weather (12 hours or 24 hours before) is a better predictor of bike usage
```{r}
#create variables representing temperature 12 and 24 hours before
lagged_bike_data = bike_data %>% 
mutate(tmp12 =lag(t1, n = 12)) %>%
mutate(tmp24 =lag(t1, n = 24)) %>% 
mutate(hum12 =lag(hum, n = 12)) %>%
mutate(hum24 =lag(hum, n = 24))

#remove rows that don't have data due to being lagged 
lagged_bike_data <- lagged_bike_data[-1:-24,]

#filter out times between 8 pm and 6 am and weekdays
dayUsage = lagged_bike_data %>% filter(time >= 6 & time <= 20)
weekendDayUsage <- dayUsage %>% filter(is_weekend == 1)

#after filtering for only weekend 
ggplot(weekendDayUsage, aes(y = cnt, x = t1, color = hum)) + geom_point()+ labs(title = "bike count versus temperature for weekend day usage")


ggplot(weekendDayUsage, aes(y = cnt, x = tmp12, color = hum12)) + geom_point()+ labs(title = "bike count versus temperature 12 hours before for weekend day usage") 


ggplot(weekendDayUsage, aes(y = cnt, x = tmp24, color = hum24)) + geom_point()+ labs(title = "bike count versus temperature 24 hours before for weekend day usage") 

#build models for weekend
weekendUsageT1 <- lm(cnt ~ t1 + hum, data = weekendDayUsage)
weekendUsage12 <- lm(cnt ~ tmp12 + hum12 , data = weekendDayUsage)
weekendUsage24 <- lm(cnt ~ tmp24 + hum24 , data = weekendDayUsage)

summary(weekendUsageT1) #adjusted R-squared = .4823
summary(weekendUsage12) #adjusted R-squared = .09851
summary(weekendUsage24) #adjusted R-squared = .3198 

mod_results1 <- data.frame(observed = weekendDayUsage$cnt, predicted = weekendUsageT1$fitted.values, residual = weekendUsageT1$residuals)

#residuals for model 1
ggplot(mod_results1, aes(y = observed, x = predicted))+ geom_point()
ggplot(mod_results1, aes(y = residual, x = predicted))+ geom_point() + geom_hline(yintercept = 0)

ggplot(mod_results1, aes(sample = residual)) + 
    geom_qq() + labs(title = "Model 1 results: Bike usage vs. Current Weather")

mod_results2 <- data.frame(observed = weekendDayUsage$cnt, predicted = weekendUsage12$fitted.values, residual = weekendUsage12$residuals)

#residuals for model 2
ggplot(mod_results2, aes(y = observed, x = predicted))+ geom_point()
ggplot(mod_results2, aes(y = residual, x = predicted))+ geom_point() + geom_hline(yintercept = 0)

ggplot(mod_results2, aes(sample = residual)) + 
    geom_qq()+ labs(title = "Model 2 results: Bike usage vs. Weather 12 hours ago")

mod_results3 <- data.frame(observed = weekendDayUsage$cnt, predicted = weekendUsage24$fitted.values, residual = weekendUsage24$residuals)

#residuals for model 3
ggplot(mod_results3, aes(y = observed, x = predicted))+ geom_point()
ggplot(mod_results3, aes(y = residual, x = predicted))+ geom_point() + geom_hline(yintercept = 0)

ggplot(mod_results3, aes(sample = residual)) + 
    geom_qq()+ labs(title = "Model 3 results: Bike usage vs. Weather 24 hours ago")



```

## Correlation of Morning Communte with Afternoon Commute
```{r}
weekdays <- bike_data %>%
  filter(is_weekend == 0)

morning_commutes <- weekdays %>%
  filter(grepl("08:00:00", timestamp))

afternoon_commutes <- weekdays %>%
  filter(grepl("17:00:00", timestamp))
afternoon_commutes <- afternoon_commutes[-c(519),] #one extra value

commutes <- data.frame(morning = morning_commutes$cnt, afternoon = afternoon_commutes$cnt)
ggplot(commutes, aes(y= afternoon, x = morning)) + geom_point() + geom_abline()

model_commutes <- lm(afternoon ~ morning , data = commutes)
summary(model_commutes)

#Adjusted R-squared: 0.4159
```

